---
title: "exploratory_data_analysis"
output: pdf_document
date: "2024-04-18"
author: "Linh Vu & Duc Nguyen"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Before you build any models, you should perform appropriate exploratory (or descriptive) analysis. This might include:

Descriptive statistics for all variables you intend to investigate. For quantitative variables, this includes: mean, standard deviation, 5-number summaries, and histograms and/or boxplots; and for categorical variables, this includes: a list of all factor levels, as well as counts and proportions within each level, and bar charts.

Exploration of the relationships between variables, both numerically and graphically. Consider not only the relationship between the response and explanatory variables, but also between two (or more) explanatory variables.

You should not focus on building statistical models at this stage.

```{r}
library(readr)
student_data <- read_csv('https://archive.ics.uci.edu/static/public/697/data.csv')
```

# Data wrangling #

Data wrangling, including joining two or more data sets into a single set, converting quantitative variables to categorical or collapsing categorical variables to ones with fewer levels, renaming variables and/or variable levels, creating new variables from existing ones

First, let's check to see if there are any missing values appear in the data. 

```{r}
missing_values <- sapply(student_data, function(x) sum(is.na(x)))
missing_values
```
We can see that there is no missing value in this data. 

Let's continue looking at the structure of this data: 

```{r}
str(student_data)
```

We can see that several variables like `Marital status`, `Nacionality`, `Mother's qualification`, `Father's qualification` are coded numerically but likely represent categories. We will convert these into categorical variables for easier analysis and interpretation.

```{r}
student_data$Marital.status <- as.factor(student_data$Marital.status)

student_data$Displaced <- as.factor(student_data$Displaced)

student_data$Daytime.evening.attendance <- as.factor(student_data$Daytime.evening.attendance)

student_data$Educational.special.needs <- as.factor(student_data$Educational.special.needs)

student_data$Tuition.fees.up.to.date <- as.factor(student_data$Tuition.fees.up.to.date)

student_data$Gender <- as.factor(student_data$Gender)

student_data$Scholarship.holder <- as.factor(student_data$Scholarship.holder)

student_data$Scholarship.holder <- as.factor(student_data$Scholarship.holder)
```

